---
title: "01: Multiple Linear Regression (Continuous Variables)"
author: "Yazan Aljerro"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
---

## 1. Project Overview
This analysis explores the relationship between morphological traits (horn size, mass) and the age of first reproduction in beetles using multiple linear regression.

## 2. Environment Setup & Data Loading
```{r setup, message=FALSE, warning=FALSE}
# Clear environment and load packages
rm(list=ls())
pacman::p_load(car, ggpubr, effects, psych, lmtest)

# Load data using relative paths
df_beetle <- read.csv(file = "../data/beetle_age_first_rep_data.csv")
```

## 3. Exploratory Data Analysis
We verify the structure of the data and check for basic descriptive statistics.
```{r eda}
# Display structure and basic stats
str(df_beetle)
describe(df_beetle$age_first_reproduction_days, type = 2)

# Check correlations
cor(df_beetle$age_first_reproduction_days, df_beetle$beetle_horn_mm)
cor(df_beetle$age_first_reproduction_days, df_beetle$mass_beetle)
```

## 4. Model Building
We construct a linear model predicting the age of first reproduction based on horn size and body mass.
```{r modeling}
mod1 <- lm(age_first_reproduction_days ~ beetle_horn_mm + mass_beetle, data = df_beetle)
summary(mod1)
```

## 5. Model Diagnostics

### 5.1 Checking Assumptions (Normality & Homoscedasticity)
```{r assumptions}
# Normality of Residuals
plot(mod1, 2)
ggqqplot(mod1$residuals)
shapiro.test(mod1$residuals)

# Homoscedasticity
plot(mod1, 1)
bptest(mod1)

# Multicollinearity 
car::vif(mod1)
```

### 5.2 Influential Cases and Outliers
```{r outliers, warning=FALSE}
# Cook's Distance
range(cooks.distance(mod1))
plot(mod1, 4)
max(cooks.distance(mod1)) > 1

# Hatvalues and DFFITS
range(hatvalues(mod1))
plot(hatvalues(mod1), type = "h")

range(dffits(mod1))
plot(dffits(mod1))

# DFBETAS
mod1_dfbetas <- dfbeta(mod1)
plot(mod1_dfbetas[,1], main="DFbetas: Intercept")
plot(mod1_dfbetas[,2], main="DFbetas: Horn Size")
plot(mod1_dfbetas[,3], main="DFbetas: Mass")
```

### 5.3 Model Comparison (ANOVA)
```{r anova_comparison}
# Creating null models for comparison
mod1_null_model_wo_beetle_horn_mm <- lm(age_first_reproduction_days ~ mass_beetle, data = df_beetle)
mod1_null_model_wo_mass_beetle <- lm(age_first_reproduction_days ~ beetle_horn_mm, data = df_beetle)

# ANOVA testing
anova(mod1_null_model_wo_beetle_horn_mm, mod1)
anova(mod1_null_model_wo_mass_beetle, mod1)

# Final Confidence Intervals
confint(object = mod1)
```

## 6. Visualization
Visualizing the main effects of the predictors on the age of first reproduction.
```{r plot, echo=TRUE, fig.cap="Main Effects of Beetle Traits"}
# Display plot in the HTML document
plot(allEffects(mod1))

# Save the plot to the plots folder using base graphics device
png(filename = "../plots/01_continuous_model.png", width = 800, height = 600, res = 120)
plot(allEffects(mod1))
dev.off() # Close the device and save the file
```