---
title: "02: Multiple Linear Regression (Categorical Variables)"
author: "Yazan Aljerro"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
---

## 1. Project Overview
This analysis evaluates the Flight Initiation Distance (FID) in animals using categorical predictors (Sex and Dominance). It demonstrates how to handle dummy coding and establish reference levels in linear regression models.

## 2. Environment Setup & Data Pre-processing
```{r setup, message=FALSE, warning=FALSE}
# Clear environment and load packages
rm(list=ls())
pacman::p_load(car, psych, lmtest, effects)

# Load data using relative paths
df_fid <- read.csv(file = "../data/fid_data.csv")

# Convert variables to factors and set reference levels
df_fid$sex <- as.factor(df_fid$sex)
df_fid$dominance <- as.factor(df_fid$dominance)

# Set baselines: Male (M) and Subordinate
df_fid$sex <- relevel(df_fid$sex, ref = "M")
df_fid$dominance <- relevel(df_fid$dominance, ref = "subordinate")

str(df_fid)
```

## 3. Exploratory Data Analysis
Visualizing the distribution of the response variable and examining the categorical predictors.
```{r eda}
# Response variable distribution
describe(df_fid$fid, type = 2)
hist(df_fid$fid, main="Distribution of Flight Initiation Distance", xlab="FID")

# Categorical predictor frequencies
plot(df_fid$sex, main="Frequency of Sex")
plot(df_fid$dominance, main="Frequency of Dominance")
table(df_fid$sex, df_fid$dominance)
```

## 4. Model Building
We construct a linear model predicting FID based on Sex and Dominance.
```{r modeling}
mod1 <- lm(fid ~ sex + dominance, data = df_fid)
summary(mod1)
```

## 5. Model Diagnostics

### 5.1 Checking Assumptions
```{r assumptions}
# Normality of Residuals
plot(mod1, 2)
shapiro.test(mod1$residuals)

# Homoscedasticity
plot(mod1, 1)
bptest(mod1)

# Multicollinearity 
vif(mod1)
```

### 5.2 Influential Cases and Outliers
```{r outliers, warning=FALSE}
# Cook's Distance
plot(mod1, 4)
max(cooks.distance(mod1)) > 1

# Hatvalues and DFFITS
plot(hatvalues(mod1), type = "h", main="Hatvalues")
plot(dffits(mod1), main="DFFITS")
max(dffits(mod1)) > 2

# DFBETAS
mod1_dfbetas <- dfbeta(mod1)
plot(mod1_dfbetas[,1], main="DFbetas: Intercept")
plot(mod1_dfbetas[,2], main="DFbetas: Sex (F)")
plot(mod1_dfbetas[,3], main="DFbetas: Dominance (Dominant)")
```

### 5.3 Model Comparison (ANOVA)
```{r anova_comparison}
# Creating null models to test the significance of categorical predictors
mod1_null_model_wo_sex <- lm(fid ~ dominance, data = df_fid)
mod1_null_model_wo_dominance <- lm(fid ~ sex, data = df_fid)

# ANOVA testing
anova(mod1_null_model_wo_sex, mod1)
anova(mod1_null_model_wo_dominance, mod1)

# Confidence Intervals
confint(object = mod1)
```

## 6. Visualization
Visualizing the main effects of the categorical predictors on FID.
```{r plot, echo=TRUE, fig.cap="Main Effects of Sex and Dominance on FID"}
# Display plot in the HTML document
plot(allEffects(mod1))

# Save the plot to the plots folder
png(filename = "../plots/02_categorical_model.png", width = 800, height = 600, res = 120)
plot(allEffects(mod1))
dev.off()
```