---
title: "03: Multiple Linear Regression (Interaction Effects)"
author: "Yazan Aljerro"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: flatly
    toc: yes
    toc_float: yes
---

## 1. Project Overview
This analysis investigates the interaction effects between categorical variables—specifically, how an animal's Sex and the Season interact to influence Flight Initiation Distance (FID). Modeling interactions is crucial for understanding complex biological dependencies.

## 2. Environment Setup & Data Pre-processing
```{r setup, message=FALSE, warning=FALSE}
# Clear environment and load packages
rm(list=ls())
pacman::p_load(car, psych, lmtest, effects)

# Load data using relative paths
df_fid_int <- read.csv(file = "../data/fid_sex_season_int.csv")

# Convert variables to factors
df_fid_int$sex <- as.factor(df_fid_int$sex)
df_fid_int$season <- as.factor(df_fid_int$season)

# Set baselines: Male (M) and Spring
df_fid_int$sex <- relevel(df_fid_int$sex, ref = "M")
df_fid_int$season <- relevel(df_fid_int$season, ref = "spring")

# Verify structure and check for missing values
str(df_fid_int)
any(is.na(df_fid_int))
```

## 3. Exploratory Data Analysis
Visualizing the distribution of the response variable and examining the cross-tabulation of our categorical predictors.
```{r eda}
# Response variable distribution
describe(df_fid_int, type = 2)
hist(df_fid_int$fid, main="Distribution of Flight Initiation Distance", xlab="FID")

# Categorical predictor frequencies and interactions
plot(df_fid_int$sex, main="Frequency of Sex")
plot(df_fid_int$season, main="Frequency of Season")
table(df_fid_int$sex, df_fid_int$season)
```

## 4. Model Building (Interaction Term)
We construct a linear model predicting FID based on Sex, Season, and their interaction (`sex * season`).
```{r modeling}
mod1 <- lm(fid ~ sex * season, data = df_fid_int)
summary(mod1)
```

## 5. Model Diagnostics

### 5.1 Checking Assumptions
```{r assumptions}
# Normality of Residuals
plot(mod1, 2)
shapiro.test(mod1$residuals)

# Homoscedasticity
plot(mod1, 1)
bptest(mod1)

# Multicollinearity 
# Note: VIF is often naturally high when interaction terms are included
vif(mod1)
```

### 5.2 Influential Cases and Outliers
```{r outliers, warning=FALSE}
# Cook's Distance
plot(mod1, 4)
max(cooks.distance(mod1)) > 1

# Hatvalues and DFFITS
plot(hatvalues(mod1), type = "h", main="Hatvalues")
plot(dffits(mod1), main="DFFITS")
max(dffits(mod1)) > 2

# DFBETAS (Evaluating impact on individual coefficients including the interaction)
mod1_dfbetas <- dfbeta(mod1)
plot(mod1_dfbetas[,1], main="DFbetas: Intercept (M, Spring)")
plot(mod1_dfbetas[,2], main="DFbetas: Sex (F)")
plot(mod1_dfbetas[,3], main="DFbetas: Season (Fall)")
plot(mod1_dfbetas[,4], main="DFbetas: Interaction (F x Fall)")
```

### 5.3 Model Comparison (ANOVA)
```{r anova_comparison}
# Creating a null model without the interaction term to test its significance
mod1_null_model_no_interaction <- lm(fid ~ sex + season, data = df_fid_int)

# ANOVA testing to see if the interaction significantly improves the model
anova(mod1_null_model_no_interaction, mod1)

# Confidence Intervals
confint(object = mod1)
```

## 6. Visualization
Visualizing the interaction effect. Non-parallel lines in the effect plot indicate a significant interaction between Sex and Season.
```{r plot, echo=TRUE, fig.cap="Interaction Effect of Sex and Season on FID"}
# Display interaction plot in the HTML document
plot(allEffects(mod1))

# Save the plot to the plots folder
png(filename = "../plots/03_interaction_plot.png", width = 800, height = 600, res = 120)
plot(allEffects(mod1))
dev.off()
```